---
output: 
  md_document:
    variant: markdown_github
---

| \-\-\-
| permalink: /research/resp%5Fexample
| title: "Remote effects spatial process (RESP) model example"
| description: "Code and output to demonstrate use of the RESP model."
| date: 2018-06-08
| toc: true
| \-\-\-


```{r setup, include=FALSE, cache=F}
knitr::opts_chunk$set(echo = T, cache = T, warning = F, message = F, 
                      fig.align = 'center', results = 'hold')
```


# At a glance

- Demonstrate the `telefit` package's implementation of the 
  [RESP model](https://arxiv.org/abs/1612.06303){:target="_blank"}.
- Reproduce figures and tables in the article linked above.
- Download the code as a [script file](../files/resp/resp_example.R).
- Download the 
  [data (resp_coex.RData)](https://figshare.com/s/f009bccc11537f46a4dc){:target="_blank"}.


# Introduction

We fit the RESP model to average monthly precipitation over Colorado in winter, 
as described [here](https://arxiv.org/abs/1612.06303){:target="_blank"}. The R 
package `telefit`
(hosted on [CRAN](http://CRAN.R-project.org/package=telefit){:target="_blank"} 
and [Github](https://github.com/jmhewitt/telefit){:target="_blank"}) 
implements the RESP model.  Basic use of the `telefit` package is demonstrated 
below.

You may need to install a few packages before proceeding and you can use the 
`devtools` package to install `telefit`.  The `telefit` package has several
dependencies and uses C++; it may take several minutes to install.


```{r install, cache=F}

# simple looping
if (!require("foreach")) install.packages("foreach")

# parallelization
if (!require("doMC")) install.packages("doMC")
if (!require("doRNG")) install.packages("doRNG")

# data manipulation and plotting
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("cowplot")) install.packages("cowplot")

# MCMC tools
if (!require("coda")) install.packages("coda")

# RESP model
if (!require("telefit")) install.packages("telefit")

# development version of RESP model
# if (!require("devtools")) install.packages("devtools")
# if (!require("telefit")) devtools::install_github('jmhewitt/telefit')
```


# Data

Load the example data 
[resp_coex.RData](https://figshare.com/s/f009bccc11537f46a4dc){:target="_blank"} 
in an R  session.  The `dat` object contains the response variable  
(precipitation) and  covariates.  Other objects contain settings and sample 
model output.  The `dat.test` and `dat.train` objects are associated with 
`fcst.test` and `fit.train`, which are used later to illustrate posterior 
predictions.

```{r load}
load('resp_coex.RData')
str(dat)
```


# Modeling

## Fit the RESP model

The RESP model may take several hours to fit because it is a hierarchical 
Bayesian spatial model.  The file 
[resp_coex.RData](https://figshare.com/s/f009bccc11537f46a4dc){:target="_blank"} 
contains sample (sub)model output `fit`, `fit.local`, `fit.remote` if you want 
to skip this step.


```{r fit, eval=F}

# sampler parameters (Note: cov.r "nugget" is not currently estimated)
maxIt = 21000
burn = 1000
priors = list(
  beta = list( Lambda = diag(10, ncol(dat$X[,,1]))), 
  cov.s = list( smoothness = .5, range = c(1, 600), variance = c(2, 30), 
                nugget = c(2, 1) ),
  cov.r = list( smoothness = .5, range = c(1, 600), variance = c(2,1e-1), 
                nugget = c(2,1) )
)

# parallelize estimation of (sub)models
ncores = detectCores() - 1
registerDoMC(ncores)

# fit (sub)models in parallel
ret = foreach(i=1:3) %dorng% {
  if(i==1) { # RESP model
    stFit(dat, priors, maxIt, coords.knots = coords.knots)
  } else if (i==2) { # SP submodel
    stFit(dat, priors, maxIt, coords.knots = coords.knots, localOnly = T)
  } else if (i==3) { # RE submodel
    stFit(dat, priors, maxIt, coords.knots = coords.knots, remoteOnly = T)
  }
}

# extract model fits
fit = ret[[1]]
fit.local = ret[[2]]
fit.remote = ret[[3]]
rm(ret)
```


The `telefit` package saves posterior samples of the RESP model's mean and  
covariance parameters.  Teleconnection effects are estimated  separately, via 
composition sampling as spatial random effects.  The posterior predictive 
distribution is also sampled via composition.  Posterior predictive samples for 
`fcst`, `fcst.local`, and `fcst.remote` are not included in 
[resp_coex.RData](https://figshare.com/s/f009bccc11537f46a4dc){:target="_blank"} 
to reduce the file's size. However, posterior sampling is relatively quick 
because it is easily parallelized.


```{r posterior_samples}

# posterior parameter samples
str(fit$parameters$samples)
```

```{r composition_sampling, eval=F}

# estimate teleconnection effects and make predictions
fcst.local = stPredict(stFit = fit.local, stData = dat, 
                       stDataNew = dat, burn = burn, ncores = ncores)
fcst.remote = stPredict(stFit = fit.remote, stData = dat, stDataNew = dat, 
                        burn = burn, ncores = ncores, returnFullAlphas = F)
fcst = stPredict(stFit = fit, stData = dat, stDataNew = dat, burn = burn, 
                 ncores = ncores, returnFullAlphas = T)
```


## Evaluate the RESP model

The `stEval()` function in `telefit` evaluates predictions with respect to 
several measures.  Variance inflation factors are also included in the 
`stVIF()` function.


```{r evaluate_predictions, results='markup'}

# evaluate predictions
clim = rowMeans(dat$Y)
fcst = stEval(fcst, dat$Y, clim)
fcst.local = stEval(fcst.local, dat$Y, clim)
fcst.remote = stEval(fcst.remote, dat$Y, clim)

# look at errors for predictions at first timepoint
str(fcst$pred[[1]]$err)

# compute VIFs
vif = stVIF(stData = dat, stFit = fit, burn = 1000)
vif$beta
summary(vif$alpha)
```


# Plots and output

## Teleconnection schematic

The `telefit` package supports basic plotting and labeling of climate data.  
Complex graphics can be built using additional functions in `ggplot2`.


```{r figure_1, fig.height=3, fig.width=8}

# unnormalize data
dat$Z = dat$Z*5252

# set plotting label
dat$Z.lab = 'z'

# set plot year
t = 1982

# build base plot of sea surface temperatures
x = plot(dat, type='remote', t=t, boxsize = 6.5) + ggtitle('')

# re-normalize data
dat$Z = dat$Z/5252

# add local precipitation
x = x + 
  geom_point(aes(x=lon, y=lat, colour = precip), inherit.aes = F, size = .6,
             stroke = 0, pch = 15,
             data = data.frame(lon = dat$coords.s[,1], lat = dat$coords.s[,2],
                               precip = dat$Y[,which(dat$tLabs==t)])) +
  scale_color_distiller('Y', palette = 'PuBuGn', direction = 1) +
  theme(legend.box = 'horizontal',
        legend.box.margin = margin(t=0, r=-8, b=0, l=0, unit='pt'))

# set left endpoints of arrows for schematic
arrows.start = data.frame(
  x = c(-100, -150, -200, -180),
  y = c(-5, 5, 35, 55)
)[c(3,2,1,4),] %>% mutate(x=ifelse(x<=0,x,x-360))

# set arrow color
col = 'forestgreen'

# plot after adding schematic arrows and labels
x + geom_curve(aes(x = x1, y = y1, xend = x2, yend = y2),
               curvature = .3, arrow = arrow(length = unit(0.03, "npc")),
               inherit.aes = F, lwd = 1, col=col,
               data = data.frame(x1=arrows.start$x[1:3],
                                 y1=arrows.start$y[1:3],
                                 x2=dat$coords.s[c(161,187,219),1],
                                 y2=dat$coords.s[c(161,187,219),2])
               ) +
  geom_curve(aes(x = x1, y = y1, xend = x2, yend = y2),
             curvature = -.3, arrow = arrow(length = unit(0.03, "npc")),
             inherit.aes = F, lwd = 1, col=col,
             data = data.frame(x1=arrows.start$x[4],
                               y1=arrows.start$y[4],
                               x2=dat$coords.s[23,1],
                               y2=dat$coords.s[23,2])
             ) +
  geom_point(aes(x=x, y=y), size=3, col=col,
             data = arrows.start, inherit.aes = F, pch=19) +
  geom_label(aes(x=x,y=y,label=label), inherit.aes = F,
             data = data.frame(
               x=c(-185,-110),
               y=c(15,dat$coords.s[94,2]+10),
               label=c('z(r,t)','x(s,t), Y(s,t)')
             ), col=col, size=5)
```


## Exploratory teleconnection analysis

The `telefit` package also supports basic exploratory data analysis
for teleconnection, including functions that compute and plot
[empirical orthogonal functions](https://climatedataguide.ucar.edu/climate-data-tools-and-analysis/empirical-orthogonal-function-eof-analysis-and-rotated-eof-analysis){:target="_blank"}
and their correlations with local response variables.

```{r figure_2, fig.width=8, fig.height=3}

plot_grid(
  plot(dat, type='eof') +
    ggtitle('SST Empirical orthogonal function 1 (EOF 1)') +
    theme(text = element_text(size=8)),
  plot(dat, type='eof_cor', signif.level = .05, signif.telecon = T,
       lwd=.7, alpha=.6) +
    ggtitle('Correlation between Precip. and EOF 1 score') +
    theme(text = element_text(size=8)),
  ncol=2,
  labels=paste(LETTERS[1:2], ')', sep='')
)
```


## RESP Teleconnection estimates

Of course, the `telefit` package plots posterior estimates of teleconnection 
effects.


```{r figure_3, fig.width=8, fig.height=3}

# build base plot of teleconnection effect estimates
resp = plot(fcst, type='eof_alpha_knots', stFit = fit, stData = dat,
            signif.telecon = T, lwd=0.7, alpha=.6) +
  scale_fill_gradient2(expression(hat(alpha)~"'"(~bold(".")~","~1)),
                       low = "#0571b0", mid = '#f7f7f7', high = '#ca0020') +
  ggtitle('Estimated teleconnection effects associated with EOF 1') 

# plot after adding denver as a point of reference
resp +
  geom_point(aes(x=lon, y=lat), 
             data = data.frame(lon = -104.9903, lat = 39.7392),
             inherit.aes = F, size = 2) + 
  geom_label(aes(x=lon, y=lat, label=label),
             data = data.frame(lon = -103.5, lat = 39.7392, label = 'Denver'),
             inherit.aes = F, alpha = .85) + 
  theme(text = element_text(size=8))
```


## Model comparison

The [resp_coex.RData](https://figshare.com/s/f009bccc11537f46a4dc){:target="_blank"} 
file contains model comparison information in the `errs` object.

```{r figure_4, fig.width=7, fig.height=3}

# set grouping labels
ilabs = c('Submodels', 'Common models')

# build plot data
errs = errs %>% mutate(model = factor(model, levels = c('RESP', 'RE', 'SP', 
                                                        'SVC', 'ENSO-T', 'CCA', 
                                                        'CLIM')),
         crps.rel = crps.cat / median(errs$crps.cat[errs$model=='CLIM']),
         comp = recode(model, 'RE'=ilabs[1], 'SP'=ilabs[1], 'RESP'='',
                       'SVC'=ilabs[2], 'ENSO-T'=ilabs[2], 'CCA'=ilabs[2], 
                       'CLIM'=ilabs[2]))

# plot error comparison
errs %>% ggplot(aes(x=model, y=crps.rel)) +
  geom_boxplot() +
  geom_hline(yintercept = 1, lty = 2) +
  xlab('') +
  ylab('Relative RPS') +
  scale_y_continuous(breaks = c(.5, 1, 1.5, 2, 3)) +
  facet_grid(~comp, scales = 'free_x', space = 'free_x') +
  theme(strip.background = element_blank(),
        strip.text = element_text(colour = 'gray15'),
        panel.border = element_rect(colour = 'gray60', linetype = 1, size = .5),
        axis.line.y = element_blank(),
        panel.spacing = unit(0, 'pt'))
```

```{r supplement_figure_c1, fig.width=7, fig.height=3}

errs %>% ggplot(aes(x=model, y=cat.heidke.alt)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, lty=2) +
  xlab('') +
  ylab('Heidke skill score') +
  coord_equal(ratio = 2) +
  facet_grid(~comp, scales = 'free_x', space = 'free_x') +
  theme(strip.background = element_blank(),
        strip.text = element_text(colour = 'gray15'),
        panel.border = element_rect(colour = 'gray60', linetype = 1, size = .5),
        axis.line.y = element_blank(),
        panel.spacing = unit(0, 'pt'))
```


## Parameter estimates

The `coda` package can easily process posterior parameter samples to develop 
point estimates and credible intervals.


```{r table_1}

maxIt = 21e3
burn = 1e3

# extract posterior samples of parameters
params = data.frame(fit$parameters$samples) %>% 
  select(-ll, -sigmasq_r_eps) %>%
  slice(burn:maxIt)
colnames(params)[1:ncol(dat$X)] = colnames(dat$X)

# "unexpand" the nugget parameter
params[,5] = params[,5] * params[,3]

# compute highest posterior density intervals
hpds = HPDinterval(mcmc(params))
hpds = round(hpds, 2)

# compute posterior estimates
ests = round(colMeans(mcmc(params)),2)

# stronger rounding for range parameters
hpds[6:7,] = round(hpds[6:7,], 0)
ests[6:7] = round(ests[6:7], 0)

# assemble parameter estimates table
df = data.frame(
  Param = c('$\\beta_0$', '$\\beta_T$', '$\\sigma^2_w$', '$\\sigma^2_\\alpha$',
            '$\\sigma^2_\\varepsilon$', '$\\rho_w$', '$\\rho_\\alpha$'),
  Est=ests,
  HPD=paste('(', paste(hpds[,1], hpds[,2], sep=', '), ')', sep='')
)

# text formatting
df$Est = gsub('-', '$-$',sprintf('%.2f', df$Est))
df[,1] = as.character(df[,1])
df[,3] = gsub('-','$-$',as.character(df[,3]))

colnames(df) = c('', 'Posterior mean', '95\\% HPD')
rownames(df) = c('Local effects',' ',  '  ', '   ',
                 'Covariance', '    ', '     ')

df
```


## Posterior predictive maps

Sample map of posterior predictions (B) and uncertainty (C), compared to truth 
(A).

```{r supplement_figure_d1, fig.width=8, fig.height=6}

plot_grid(
  plot(dat.test, fill.lab=expression(Y)) + ggtitle('Observed (1982)'),
  plot(fcst, fill.lab=expression(hat(Y))) + ggtitle('Forecast') +
    theme(axis.title.y = element_text(color='white')),
  NULL,
  plot(fcst, type='se', fill.lab=expression(SE(hat(Y)))) + 
    ggtitle('Uncertainty'),
  ncol=2, labels = c('A)','B)', '', 'C)')
)
```


Similar maps for discretized posterior predictions.

```{r supplement_figure_d2, fig.width=8, fig.height=6}

# compute posterior logits
logits = qlogis(fcst.test$samples$cat_probs)
logits.est = matrix(nrow = nrow(logits), ncol = ncol(logits))
for(i in 1:nrow(logits)) {
  logits.est[i,] = rowMeans(logits[i,,])
}

# transfer logits to a plottable object
logitDat = dat.test
logitDat$Y = apply(logits.est, 1, max) 

# build plots

p = plot(dat.test, type='cat.response', 
         category.breaks = fcst$category.breaks, fill.lab='') +
  ggtitle('Observed (1982)')

uncty = plot(logitDat, fill.lab=expression(hat(logit)(tilde(Y)))) +
  ggtitle('Uncertainty')

# assemble and display plots
ggdraw(plot_grid(
  plot_grid(
    plot_grid(
      p + theme(legend.position = 'none'),
      NULL,
      plot(fcst.test, type='cat.pred', fill.lab='') +
        ggtitle('Forecast') +
        theme(axis.title.y = element_text(color='white'),
              legend.position = 'none'),
      nrow=1, align='v', rel_widths = c(1,.03,1), labels=c('A)','','B)')),
    get_legend(p),
  rel_widths=c(1, 0.2)
  ),
    plot_grid(
    plot_grid(
      NULL,
      NULL,
      uncty + theme(legend.position = 'none'),
      nrow=1, align='v', rel_widths = c(1,.03,1), labels=c('','','C)')),
    get_legend(uncty),
  rel_widths=c(1, 0.2)
  ),
  nrow=2
))
```


## Knot locations

For completeness, we can view the knot locations used during estimation.

```{r supplement_figure_d3, fig.width=8, fig.height=3}

dat$Z.lab = 'SST'
plot(dat, type='remote', coords.knots = coords.knots, t = 1982) +
  ggtitle('')
```